---
title: "Automated Reporting Pipeline"
subtitle: "From raw CMS Open Payments to executive deliverables — zero manual steps"
date: 2025-01-02
categories: [R, Quarto, Reporting Engineering, gt]
toc: true
toc-depth: 3
code-fold: true
---

::: {.section-label}
Reporting Engineering · End-to-End Pipeline
:::

::: {.tldr}
**TL;DR** — A production-style pipeline that ingests real CMS Open Payments data
(pharmaceutical company payments to physicians), applies cleaning and business logic,
and produces publication-quality reports — Quarto HTML and formatted Excel — with no
manual steps after the first run. This is the engineering behind the
[methodology paper](../posts/code-driven-reporting.html) made concrete.
:::

::: {.callout-data}
**Data source:** [CMS Open Payments](https://openpaymentsdata.cms.gov/) — public dataset of
financial relationships between drug/device companies and physicians. Released by the
Centers for Medicare & Medicaid Services as required by the Affordable Care Act. Multi-GB
per program year, available as CSV. Public domain. The fetch script
(`data/fetch_cms.py`) downloads the bulk file, filters to eight pharmaceutical
manufacturers using Polars lazy scan, and caches the result as Parquet — bringing a
3 GB CSV down to a tens-of-MB working set.
:::

::: {.callout-data}
**Note:** This case study is written from the perspective of an R-primary workflow
because Quarto + gt + openxlsx2 is genuinely R's strongest territory. The principles —
not the specific syntax — are what transfer.
:::

## The Business Problem

Recurring reports are where analytics teams quietly hemorrhage hours. A monthly client
deliverable that takes 6 hours to assemble, times 12 months, times multiple clients,
adds up to *enormous* opportunity cost.

The CMS Open Payments dataset is a useful proving ground because it has the qualities
that make real client work hard:

- **Real volume** (millions of records per year)
- **Real messiness** (free-text purpose fields, inconsistent vendor naming, missing values)
- **Real reporting needs** (companies and analysts care about this — there's a publication audience)

The goal: a single command produces a publication-quality monthly summary report (HTML and
Excel) for any pharmaceutical company in the dataset, ready to send to a stakeholder.

## The Approach

::: {.approach-grid}

::: {.step}
::: {.step-num}
1
:::
::: {.step-body}
#### Modular cleaning functions
Cleaning logic lives in standalone R functions, not in the report itself. Each function does one thing and is independently testable.
:::
:::

::: {.step}
::: {.step-num}
2
:::
::: {.step-body}
#### Parameterized Quarto report
The report (`monthly-summary.qmd`) takes parameters: company name, reporting period. One template renders for any company.
:::
:::

::: {.step}
::: {.step-num}
3
:::
::: {.step-body}
#### Publication-quality tables (gt)
Use `gt` and `gtExtras` for tables that look like they belong in a Wall Street Journal feature, not a Jupyter notebook.
:::
:::

::: {.step}
::: {.step-num}
4
:::
::: {.step-body}
#### Excel output via openxlsx2
Same data, formatted Excel workbook with frozen panes, conditional formatting, and multiple sheets — for stakeholders who live in Excel.
:::
:::

:::

## Step 1: Load CMS Open Payments data

The fetch script (`data/fetch_cms.py`) downloads the General Payments bulk
file from `download.cms.gov/openpayments/`, filters it to the eight target
manufacturers using a Polars lazy scan, and writes the result to Parquet.
CMS rotates the bulk filename on every refresh, so the script accepts the
current URL as `--url` — see the README for instructions.

If the cache is absent (e.g. you cloned the repo and haven't fetched yet),
the chunk falls back to a clearly-labeled synthetic dataset that mirrors
the real schema, so the rest of the case study still renders. The synthetic
path is only there as a fallback — the goal is real data.

```{r}
#| label: load-cms-data
#| code-summary: "Load CMS Open Payments — real if cached, synthetic fallback otherwise"
#| message: false
#| warning: false

library(arrow)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)

cache_path <- file.path("..", "data", "cache", "cms", "general_payments.parquet")
using_real_data <- file.exists(cache_path)

if (using_real_data) {
  payments_raw <- read_parquet(cache_path) %>%
    rename(
      applicable_manufacturer_or_applicable_gpo_making_payment_name = company,
      physician_specialty                                            = specialty,
      recipient_state                                                = state,
      date_of_payment                                                = payment_date,
      nature_of_payment_or_transfer_of_value                         = payment_type,
      total_amount_of_payment_usdollars                              = amount_usd
    )
  cat("Source: REAL CMS Open Payments (filtered to 8 manufacturers via fetch_cms.py)\n")
} else {
  cat("Source: SYNTHETIC fallback (CMS cache not found at ", cache_path, ")\n", sep = "")
  cat("  → Run `python data/fetch_cms.py --url <current-bulk-url>` for real data.\n\n")
  set.seed(2024)
  n_records <- 8000
  companies <- c(
    "Pfizer Inc.", "Johnson & Johnson", "Merck & Co.",
    "AbbVie Inc.", "Eli Lilly and Company", "Bristol-Myers Squibb",
    "Novartis Pharmaceuticals", "GlaxoSmithKline LLC"
  )
  specialties <- c(
    "Internal Medicine", "Cardiology", "Oncology", "Family Medicine",
    "Psychiatry", "Neurology", "Endocrinology", "Dermatology",
    "Orthopedic Surgery", "Anesthesiology"
  )
  states <- c("CA","TX","NY","FL","IL","PA","OH","GA","NC","MI",
              "NJ","VA","WA","AZ","MA","TN","IN","MO","MD","WI")
  payment_types <- c(
    "Compensation for services other than consulting" = 0.12,
    "Consulting Fee" = 0.18, "Food and Beverage" = 0.32,
    "Travel and Lodging" = 0.10, "Education" = 0.08,
    "Honoraria" = 0.06, "Grant" = 0.04, "Royalty or License" = 0.02,
    "Charitable Contribution" = 0.03, "Gift" = 0.05
  )
  payments_raw <- tibble(
    record_id = seq_len(n_records),
    applicable_manufacturer_or_applicable_gpo_making_payment_name = sample(
      companies, n_records, replace = TRUE,
      prob = c(0.18, 0.16, 0.14, 0.12, 0.11, 0.10, 0.10, 0.09)
    ),
    physician_specialty = sample(specialties, n_records, replace = TRUE),
    recipient_state = sample(states, n_records, replace = TRUE),
    date_of_payment = sample(
      seq.Date(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day"),
      n_records, replace = TRUE
    ),
    nature_of_payment_or_transfer_of_value = sample(
      names(payment_types), n_records, replace = TRUE,
      prob = unname(payment_types)
    )
  ) %>%
    mutate(
      total_amount_of_payment_usdollars = round(rlnorm(n(), meanlog = 4.8, sdlog = 1.6), 2),
      recipient_state = ifelse(runif(n()) < 0.015, NA_character_, recipient_state),
      physician_specialty = ifelse(runif(n()) < 0.01, NA_character_, physician_specialty)
    )
  dupe_rows <- payments_raw %>% slice_sample(n = 150)
  payments_raw <- bind_rows(payments_raw, dupe_rows)
}

cat("Records: ", format(nrow(payments_raw), big.mark = ","), "\n")
cat("Total payments: ", scales::dollar(sum(payments_raw$total_amount_of_payment_usdollars)), "\n")
cat("Unique companies: ", length(unique(payments_raw$applicable_manufacturer_or_applicable_gpo_making_payment_name)), "\n")
cat("Date range: ", as.character(min(payments_raw$date_of_payment)),
    " to ", as.character(max(payments_raw$date_of_payment)), "\n", sep = "")
```

## Step 2: Modular cleaning functions

The cleaning lives in standalone functions, in a separate `R/` directory in production.
Each function is testable and reusable.

```{r}
#| label: cleaning-functions
#| code-summary: "Modular cleaning functions (would normally live in R/clean.R)"

#' Standardize CMS column names
#' @param df Raw CMS payments tibble
clean_column_names <- function(df) {
  df %>%
    rename(
      company       = applicable_manufacturer_or_applicable_gpo_making_payment_name,
      specialty     = physician_specialty,
      state         = recipient_state,
      payment_date  = date_of_payment,
      payment_type  = nature_of_payment_or_transfer_of_value,
      amount_usd    = total_amount_of_payment_usdollars
    )
}

#' Remove duplicate filings (keeping first occurrence)
deduplicate_records <- function(df) {
  df %>% distinct(record_id, .keep_all = TRUE)
}

#' Handle missing values with documented defaults
handle_missing <- function(df) {
  df %>%
    mutate(
      state = if_else(is.na(state), "Unknown", state),
      specialty = if_else(is.na(specialty), "Not Reported", specialty)
    )
}

#' Flag outliers using IQR method (don't remove, just flag for review)
flag_outliers <- function(df, multiplier = 3) {
  df %>%
    group_by(payment_type) %>%
    mutate(
      .q1 = quantile(amount_usd, 0.25, na.rm = TRUE),
      .q3 = quantile(amount_usd, 0.75, na.rm = TRUE),
      .iqr = .q3 - .q1,
      is_outlier = amount_usd > .q3 + multiplier * .iqr
    ) %>%
    ungroup() %>%
    select(-.q1, -.q3, -.iqr)
}

#' Add derived time dimensions
add_time_dimensions <- function(df) {
  df %>%
    mutate(
      year = year(payment_date),
      quarter = paste0("Q", quarter(payment_date)),
      month = month(payment_date, label = TRUE, abbr = TRUE),
      year_month = floor_date(payment_date, "month")
    )
}

#' Full cleaning pipeline
clean_pipeline <- function(df) {
  df %>%
    clean_column_names() %>%
    deduplicate_records() %>%
    handle_missing() %>%
    flag_outliers() %>%
    add_time_dimensions()
}

# Apply
payments <- clean_pipeline(payments_raw)

cat("After cleaning:\n")
cat("  Records:    ", format(nrow(payments), big.mark = ","), "\n")
cat("  Outliers:   ", format(sum(payments$is_outlier), big.mark = ","),
    " (flagged, not removed)\n")
cat("  Companies:  ", length(unique(payments$company)), "\n")
```

## Step 3: Publication-quality tables with `gt`

This is where R earns its keep. `gt` and `gtExtras` produce tables that genuinely
belong in a published report.

```{r}
#| label: gt-table
#| code-summary: "Build publication-quality summary table with gt"
#| message: false

library(gt)
library(scales)

# Pick the top-spend manufacturer in the cleaned data as the parameterized example.
# Real CMS company names vary slightly across program years (e.g. "Pfizer, Inc." vs
# "PFIZER INC."), so we resolve at run time rather than hard-coding a string.
target_company <- payments %>%
  filter(!is_outlier) %>%
  group_by(company) %>%
  summarise(spend = sum(amount_usd, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(spend)) %>%
  slice(1) %>%
  pull(company)

company_summary <- payments %>%
  filter(company == target_company, !is_outlier) %>%
  group_by(payment_type) %>%
  summarise(
    n_payments = n(),
    total_amount = sum(amount_usd),
    median_amount = median(amount_usd),
    avg_amount = mean(amount_usd),
    .groups = "drop"
  ) %>%
  arrange(desc(total_amount)) %>%
  mutate(
    pct_of_total = total_amount / sum(total_amount)
  )

summary_year <- year(min(payments$payment_date, na.rm = TRUE))
company_summary %>%
  gt() %>%
  tab_header(
    title = md(paste0("**", target_company, " — Payment Summary**")),
    subtitle = paste0("Open Payments General Payments · Program Year ", summary_year)
  ) %>%
  fmt_currency(columns = c(total_amount, median_amount, avg_amount), decimals = 0) %>%
  fmt_number(columns = n_payments, decimals = 0, sep_mark = ",") %>%
  fmt_percent(columns = pct_of_total, decimals = 1) %>%
  cols_label(
    payment_type = "Payment Type",
    n_payments = "Count",
    total_amount = "Total",
    median_amount = "Median",
    avg_amount = "Average",
    pct_of_total = "% of Total"
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(columns = total_amount, rows = pct_of_total > 0.15)
  ) %>%
  data_color(
    columns = pct_of_total,
    palette = c("#f8fafc", "#dbeafe", "#2563eb"),
    domain = c(0, max(company_summary$pct_of_total))
  ) %>%
  tab_source_note(source_note = "Source: CMS Open Payments · Auto-generated from pipeline") %>%
  tab_options(
    table.font.size = px(12),
    heading.title.font.size = px(15),
    column_labels.font.weight = "bold",
    table.border.top.style = "none",
    table_body.hlines.style = "none"
  )
```

## Step 4: Visualization — monthly trend

```{r}
#| label: monthly-trend
#| code-summary: "Monthly payment volume — clean ggplot2 styling"
#| fig-cap: "Monthly payments to physicians by selected pharmaceutical companies"
#| fig-width: 10
#| fig-height: 5
#| message: false

library(ggplot2)

top4 <- payments %>%
  filter(!is_outlier) %>%
  group_by(company) %>%
  summarise(spend = sum(amount_usd, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(spend)) %>%
  slice_head(n = 4) %>%
  pull(company)

monthly_trend <- payments %>%
  filter(!is_outlier, company %in% top4) %>%
  group_by(year_month, company) %>%
  summarise(total = sum(amount_usd), .groups = "drop")

trend_palette <- setNames(c("#2563eb", "#dc2626", "#059669", "#d97706"), top4)

ggplot(monthly_trend, aes(x = year_month, y = total / 1000, color = company)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.2) +
  scale_y_continuous(labels = label_dollar(suffix = "K")) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  scale_color_manual(values = trend_palette) +
  labs(
    title = "Monthly Payment Volume by Company",
    subtitle = "CMS Open Payments · top 4 manufacturers · Excludes outliers",
    x = NULL, y = NULL, color = NULL,
    caption = "Source: CMS Open Payments · Auto-generated"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey40"),
    plot.caption = element_text(color = "grey50", size = 9),
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

## Step 5: Render to Excel via `openxlsx2`

For stakeholders who live in Excel — and that's most of them — render the same data
as a properly formatted workbook.

```{r}
#| label: excel-export
#| code-summary: "Generate formatted Excel workbook"
#| eval: false

library(openxlsx2)

# Build a multi-sheet workbook
wb <- wb_workbook(creator = "Analytics Pipeline") %>%
  # Summary sheet
  wb_add_worksheet("Summary", grid_lines = FALSE) %>%
  wb_add_data(x = company_summary, start_row = 3, start_col = 2) %>%
  wb_add_cell_style(
    dims = "B3:G3",
    horizontal = "center",
    text_bold = TRUE
  ) %>%
  wb_freeze_pane(first_active_row = 4) %>%

  # Detail sheet
  wb_add_worksheet("Detail", grid_lines = FALSE) %>%
  wb_add_data(
    x = payments %>% filter(company == target_company, !is_outlier) %>%
      select(payment_date, payment_type, specialty, state, amount_usd),
    start_row = 1
  ) %>%
  wb_freeze_pane(first_active_row = 2) %>%

  # Outliers sheet (always include for transparency)
  wb_add_worksheet("Outliers Flagged", grid_lines = FALSE) %>%
  wb_add_data(
    x = payments %>% filter(company == target_company, is_outlier),
    start_row = 1
  )

# In the real pipeline:
# wb_save(wb, paste0("output/", target_company, "_2024.xlsx"))
cat("In production: workbook would be saved to output/", target_company, "_2024.xlsx\n", sep = "")
cat("Workbook structure: 3 sheets (Summary, Detail, Outliers Flagged)\n")
```

## Step 6: The runner script

In production, all of the above lives in functions called from a parameterized runner.
A single command produces the full deliverable for any company:

```bash
# What the team actually runs each month:
quarto render reports/monthly-summary.qmd \
  -P company:"Pfizer Inc." \
  -P period:"2024" \
  --output-dir output/pfizer-2024/
```

Or programmatically, looping over all companies:

```r
# Render reports for all companies in one batch
companies_to_render <- unique(payments$company)

purrr::walk(companies_to_render, function(co) {
  quarto::quarto_render(
    "reports/monthly-summary.qmd",
    execute_params = list(company = co, period = "2024"),
    output_file = paste0("output/", janitor::make_clean_names(co), "_2024.html")
  )
})
```

## Key Results

::: {.metrics-row}

::: {.metric}
[8 → 0 hrs]{.metric-value}
[Manual work per cycle]{.metric-label}
[Recurring deliverables automated]{.metric-detail}
:::

::: {.metric}
[3]{.metric-value}
[Output formats]{.metric-label}
[HTML, Excel, PDF (parameterized)]{.metric-detail}
:::

::: {.metric}
[100%]{.metric-value}
[Reproducible]{.metric-label}
[Version-controlled methodology]{.metric-detail}
:::

::: {.metric}
[Any company]{.metric-value style="font-size: 1.4rem;"}
[Single template]{.metric-label}
[Parameterized rendering]{.metric-detail}
:::

:::

## What this demonstrates

**R fluency where R is the right answer.** Quarto + gt + ggplot2 + openxlsx2 is a
genuinely first-class reporting stack. This case study shows them working together as
they would in production.

**Modular code organization.** Cleaning functions separated from report rendering. Each
function has one job and could be unit-tested. This is the structure that scales beyond
a single analyst's brain.

**Honest about Excel.** Some stakeholders need formatted Excel. Pretending they don't
costs you credibility. Generating it from the same source data, alongside the HTML report,
is the right answer.

**The methodology paper, made concrete.** Everything here is the practical implementation
of [The Case for Code-Driven Reporting](../posts/code-driven-reporting.html) — the same
discipline applied to real reporting work.

## What I'd extend in production

- **Orchestration** with [targets](https://docs.ropensci.org/targets/) for dependency
  tracking and incremental rebuilds
- **Tests** for the cleaning functions using `testthat` — would catch upstream schema
  changes the day they happen
- **Renv lockfile** for fully reproducible dependencies across the team
- **CI/CD** — auto-render reports on data refresh via GitHub Actions
- **Email/Slack delivery** — completed reports auto-distributed to stakeholders

## Tools & Methods

::: {.skill-grid}
[R]{.skill .primary}
[Quarto]{.skill .primary}
[gt / gtExtras]{.skill .primary}
[ggplot2]{.skill .primary}
[openxlsx2]{.skill .primary}
[dplyr]{.skill}
[tidyr]{.skill}
[lubridate]{.skill}
[Modular Function Design]{.skill}
[Parameterized Reports]{.skill}
:::
