---
title: "Automated Reporting Pipeline"
subtitle: "Raw CMS Open Payments to finished deliverables, zero manual steps after the first run"
date: 2026-05-31
categories: [R, Quarto, Reporting Engineering, gt]
toc: true
toc-depth: 3
code-fold: true
---

::: {.section-label}
Reporting Engineering · End-to-End Pipeline
:::

::: {.tldr}
**TL;DR** — A production style pipeline that ingests real CMS Open Payments data
(pharma company payments to physicians), runs it through modular cleaning functions,
and produces publication quality reports in Quarto HTML and formatted Excel with no
manual steps after the first run. This is the engineering behind the
[methodology paper](../posts/code-driven-reporting.html) made concrete, on 3.8 million
real records.
:::

::: {.callout-data}
**Data source:** [CMS Open Payments](https://openpaymentsdata.cms.gov/), the public dataset of
financial relationships between drug and device companies and physicians, released by the
Centers for Medicare & Medicaid Services under the Affordable Care Act. The bulk file runs
multiple GB per program year. The fetch script (`data/fetch_cms.py`) downloads it, filters to
eight pharmaceutical manufacturers with a Polars lazy scan, and caches the result as Parquet.
A 3 GB CSV becomes a 41 MB working set.
:::

::: {.callout-data}
**Note:** this case study is written from an R-first perspective because Quarto + gt +
openxlsx2 is genuinely R's strongest territory. The principles transfer, the syntax is
just the costume.
:::

## The Business Problem

Recurring reports are where analytics teams quietly bleed hours. A monthly client
deliverable that takes 6 hours to assemble, times 12 months, times multiple clients,
that adds up to real money and nobody budgets for it.

The CMS Open Payments dataset is a useful proving ground because it has the same
qualities that make real client work hard:

- **Real volume**, millions of records per program year
- **Real messiness**, the same parent company files under a pile of legal entity names, payment type labels run to 140 characters
- **Real reporting needs**, journalists and compliance teams actually publish off this data

The goal: one command produces a publication quality summary report, HTML and Excel,
for any manufacturer in the dataset, ready to send to a stakeholder.

## The Approach

::: {.approach-grid}

::: {.step}
::: {.step-num}
1
:::
::: {.step-body}
#### Modular cleaning functions
Cleaning logic lives in standalone R functions, not buried in the report. Each function does one thing and is independently testable.
:::
:::

::: {.step}
::: {.step-num}
2
:::
::: {.step-body}
#### Parameterized Quarto report
The report template takes parameters, company name and reporting period. One template renders for any company.
:::
:::

::: {.step}
::: {.step-num}
3
:::
::: {.step-body}
#### Publication quality tables (gt)
`gt` and `gtExtras` for tables that look like they belong in a Wall Street Journal feature, not a notebook dump.
:::
:::

::: {.step}
::: {.step-num}
4
:::
::: {.step-body}
#### Excel output via openxlsx2
Same data, formatted Excel workbook with frozen panes and multiple sheets, for the stakeholders who live in Excel. Which is most of them.
:::
:::

:::

## Step 1: Load CMS Open Payments data

The fetch script (`data/fetch_cms.py`) downloads the General Payments bulk file from
`download.cms.gov/openpayments/`, filters it to the eight target manufacturers with a
Polars lazy scan, and writes Parquet. CMS rotates the bulk filename on every refresh so
the script takes the current URL as `--url`, the README has the details.

If the cache is missing (you cloned the repo and haven't fetched yet) the chunk falls
back to a clearly labeled synthetic dataset with the same schema so the page still
renders. The synthetic path is only a fallback, the goal is real data and everything
below was rendered against it.

```{r}
#| label: load-cms-data
#| code-summary: "Load CMS Open Payments — real if cached, synthetic fallback otherwise"
#| message: false
#| warning: false

library(arrow)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)

cache_path <- file.path("..", "data", "cache", "cms", "general_payments.parquet")
using_real_data <- file.exists(cache_path)

if (using_real_data) {
  ## the cache already uses short friendly names. rename them BACK to the
  ## real CMS bulk header names so the cleaning functions below face the
  ## same schema the raw file throws at you in production
  payments_raw <- read_parquet(cache_path) %>%
    rename(
      applicable_manufacturer_or_applicable_gpo_making_payment_name = company,
      physician_specialty                                            = specialty,
      recipient_state                                                = state,
      date_of_payment                                                = payment_date,
      nature_of_payment_or_transfer_of_value                         = payment_type,
      total_amount_of_payment_usdollars                              = amount_usd
    )
  cat("source: REAL CMS Open Payments (filtered to 8 manufacturers via fetch_cms.py)\n")
} else {
  cat("source: SYNTHETIC fallback (CMS cache not found at ", cache_path, ")\n", sep = "")
  cat("  -> run `python data/fetch_cms.py --url <current-bulk-url>` for real data.\n\n")
  ## same schema as the real file, obviously fake numbers
  set.seed(2024)
  n_records <- 8000
  companies <- c(
    "Pfizer Inc.", "Johnson & Johnson", "Merck & Co.",
    "AbbVie Inc.", "Eli Lilly and Company", "Bristol-Myers Squibb",
    "Novartis Pharmaceuticals", "GlaxoSmithKline LLC"
  )
  specialties <- c(
    "Internal Medicine", "Cardiology", "Oncology", "Family Medicine",
    "Psychiatry", "Neurology", "Endocrinology", "Dermatology",
    "Orthopedic Surgery", "Anesthesiology"
  )
  states <- c("CA","TX","NY","FL","IL","PA","OH","GA","NC","MI",
              "NJ","VA","WA","AZ","MA","TN","IN","MO","MD","WI")
  payment_types <- c(
    "Compensation for services other than consulting" = 0.12,
    "Consulting Fee" = 0.18, "Food and Beverage" = 0.32,
    "Travel and Lodging" = 0.10, "Education" = 0.08,
    "Honoraria" = 0.06, "Grant" = 0.04, "Royalty or License" = 0.02,
    "Charitable Contribution" = 0.03, "Gift" = 0.05
  )
  payments_raw <- tibble(
    record_id = seq_len(n_records),
    applicable_manufacturer_or_applicable_gpo_making_payment_name = sample(
      companies, n_records, replace = TRUE,
      prob = c(0.18, 0.16, 0.14, 0.12, 0.11, 0.10, 0.10, 0.09)
    ),
    physician_specialty = sample(specialties, n_records, replace = TRUE),
    recipient_state = sample(states, n_records, replace = TRUE),
    date_of_payment = sample(
      seq.Date(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day"),
      n_records, replace = TRUE
    ),
    nature_of_payment_or_transfer_of_value = sample(
      names(payment_types), n_records, replace = TRUE,
      prob = unname(payment_types)
    )
  ) %>%
    mutate(
      total_amount_of_payment_usdollars = round(rlnorm(n(), meanlog = 4.8, sdlog = 1.6), 2),
      recipient_state = ifelse(runif(n()) < 0.015, NA_character_, recipient_state),
      physician_specialty = ifelse(runif(n()) < 0.01, NA_character_, physician_specialty)
    )
  ### sprinkle in duplicate filings so the dedupe step has something to do
  dupe_rows <- payments_raw %>% slice_sample(n = 150)
  payments_raw <- bind_rows(payments_raw, dupe_rows)
}

cat("records:    ", format(nrow(payments_raw), big.mark = ","), "\n")
cat("total paid: ", scales::dollar(sum(payments_raw$total_amount_of_payment_usdollars)), "\n")
cat("raw company names:", length(unique(payments_raw$applicable_manufacturer_or_applicable_gpo_making_payment_name)), "\n")
cat("date range: ", as.character(min(payments_raw$date_of_payment)),
    " to ", as.character(max(payments_raw$date_of_payment)), "\n", sep = "")

## quick honesty check on the storage story. write the exact same rows back
## out as csv, measure both files, delete the csv. takes about two seconds
## and the key results section reads this number at render time
if (using_real_data) {
  tmp_csv <- tempfile(fileext = ".csv")
  write_csv_arrow(payments_raw, tmp_csv)
  compression_ratio <- file.size(tmp_csv) / file.size(cache_path)
  cat(sprintf("on disk: %.0f MB as csv vs %.0f MB as parquet, %.1fx smaller\n",
      file.size(tmp_csv) / 1024^2, file.size(cache_path) / 1024^2, compression_ratio))
  unlink(tmp_csv)
} else {
  compression_ratio <- NA
}
```

## Step 2: Modular cleaning functions

The cleaning lives in standalone functions, in a separate `R/` directory in production.
Each one is testable on its own and the report just calls the pipeline.

The one worth a second look is `roll_up_company()`. The raw file lists 26 different
payer names for what a stakeholder would call 7 companies. "PFIZER INC." and "PFIZER
INTERNATIONAL LLC" are not two companies, and Janssen is just Johnson & Johnson wearing
a different badge. Skip this step and every company level number in the report is wrong.

```{r}
#| label: cleaning-functions
#| code-summary: "Modular cleaning functions (would live in R/clean.R in production)"

## rename the wall-of-text CMS bulk columns to names a human can type
clean_column_names <- function(df) {
  df %>%
    rename(
      company       = applicable_manufacturer_or_applicable_gpo_making_payment_name,
      specialty     = physician_specialty,
      state         = recipient_state,
      payment_date  = date_of_payment,
      payment_type  = nature_of_payment_or_transfer_of_value,
      amount_usd    = total_amount_of_payment_usdollars
    )
}

## drop duplicate filings, keep the first occurrence
deduplicate_records <- function(df) {
  df %>% distinct(record_id, .keep_all = TRUE)
}

## missing values get documented defaults instead of silent NA drops
handle_missing <- function(df) {
  df %>%
    mutate(
      state = if_else(is.na(state), "Unknown", state),
      specialty = if_else(is.na(specialty), "Not Reported", specialty)
    )
}

## the same parent files under a pile of legal entities. roll them up so
## company level numbers mean what a reader thinks they mean
### same entity resolution problem as the FEC vendor names in the political
### spend study, just smaller. 26 names here, so a checked-by-hand regex
### map is enough. at vendor scale you graduate to fuzzy matching
roll_up_company <- function(df) {
  df %>%
    mutate(parent = case_when(
      str_detect(str_to_upper(company), "ABBVIE")          ~ "AbbVie",
      str_detect(str_to_upper(company), "PFIZER")          ~ "Pfizer",
      str_detect(str_to_upper(company), "NOVARTIS")        ~ "Novartis",
      str_detect(str_to_upper(company), "GLAXO|GSK")       ~ "GSK",
      str_detect(str_to_upper(company), "JANSSEN|JOHNSON") ~ "Johnson & Johnson",
      str_detect(str_to_upper(company), "MERCK")           ~ "Merck",
      str_detect(str_to_upper(company), "LILLY")           ~ "Eli Lilly",
      str_detect(str_to_upper(company), "BRISTOL")         ~ "Bristol Myers Squibb",
      TRUE ~ company   ## anything unmapped keeps its raw name so it's visible
    ))
}

## CMS payment type labels run up to 140 characters. nobody wants that in
## a table row, and the longest one is also the biggest dollar bucket
shorten_payment_types <- function(df) {
  df %>%
    mutate(payment_type_short = case_when(
      str_detect(payment_type, "^Compensation for services other than consulting") ~ "Speaker / services fees",
      str_detect(payment_type, "^Compensation for serving as faculty")             ~ "CME speaker fees",
      payment_type == "Food and Beverage"   ~ "Meals",
      payment_type == "Consulting Fee"      ~ "Consulting",
      payment_type == "Travel and Lodging"  ~ "Travel and lodging",
      payment_type == "Royalty or License"  ~ "Royalties",
      str_detect(payment_type, "^Space rental")            ~ "Facility fees",
      str_detect(payment_type, "^Long term medical supply") ~ "Device loans",
      TRUE ~ payment_type
    ))
}

## flag outliers with 3x IQR inside each payment type. flag, never remove.
### a $22,000 royalty is normal, a $22,000 meal is not, that's why the
### threshold is computed per payment type and not globally
flag_outliers <- function(df, multiplier = 3) {
  df %>%
    group_by(payment_type) %>%
    mutate(
      .q1 = quantile(amount_usd, 0.25, na.rm = TRUE),
      .q3 = quantile(amount_usd, 0.75, na.rm = TRUE),
      is_outlier = amount_usd > .q3 + multiplier * (.q3 - .q1)
    ) %>%
    ungroup() %>%
    select(-.q1, -.q3)
}

## derived time dimensions, every report downstream wants these
add_time_dimensions <- function(df) {
  df %>%
    mutate(
      year = year(payment_date),
      quarter = paste0("Q", quarter(payment_date)),
      month = month(payment_date, label = TRUE, abbr = TRUE),
      year_month = floor_date(payment_date, "month")
    )
}

## the full pipeline reads like the run sheet
clean_pipeline <- function(df) {
  df %>%
    clean_column_names() %>%
    deduplicate_records() %>%
    handle_missing() %>%
    roll_up_company() %>%
    shorten_payment_types() %>%
    flag_outliers() %>%
    add_time_dimensions()
}

payments <- clean_pipeline(payments_raw)

## the key results section reads these at render time, no hand-typed numbers
n_raw_names <- n_distinct(payments$company)
n_parents   <- n_distinct(payments$parent)
n_dupes     <- nrow(payments_raw) - nrow(payments)
pct_outliers <- mean(payments$is_outlier) * 100

fmt_n <- function(n) if (n >= 1e6) sprintf("%.1fM", n / 1e6) else format(n, big.mark = ",")

cat("after cleaning:\n")
cat("  records:   ", format(nrow(payments), big.mark = ","), "\n")
cat("  dupes out: ", format(n_dupes, big.mark = ","), "\n")
cat("  outliers:  ", format(sum(payments$is_outlier), big.mark = ","),
    sprintf(" (%.1f%%, flagged not removed)\n", pct_outliers))
cat("  companies: ", n_raw_names, "raw filing names ->", n_parents, "parents\n")
```

## Step 3: Publication quality tables with `gt`

This is where R earns its keep. `gt` produces tables that genuinely belong in a
published report. The summary excludes flagged outliers on purpose, they get their own
sheet in the Excel workbook later so nothing disappears, it just gets labeled.

```{r}
#| label: gt-table
#| code-summary: "Build publication-quality summary table with gt"
#| message: false

library(gt)
library(scales)

## top spend parent in the cleaned data becomes the parameterized example.
## resolved at run time, not hard coded, because the ranking is allowed to
## change when the data refreshes
target_company <- payments %>%
  filter(!is_outlier) %>%
  count(parent, wt = amount_usd, name = "spend") %>%
  slice_max(spend, n = 1) %>%
  pull(parent)

company_summary <- payments %>%
  filter(parent == target_company, !is_outlier) %>%
  group_by(payment_type_short) %>%
  summarise(
    n_payments = n(),
    total_amount = sum(amount_usd),
    median_amount = median(amount_usd),
    avg_amount = mean(amount_usd),
    .groups = "drop"
  ) %>%
  arrange(desc(total_amount)) %>%
  mutate(pct_of_total = total_amount / sum(total_amount))

summary_year <- year(min(payments$payment_date, na.rm = TRUE))
company_summary %>%
  gt() %>%
  tab_header(
    title = md(paste0("**", target_company, " — Payment Summary**")),
    subtitle = paste0("Open Payments General Payments · Program Year ", summary_year,
                      " · outliers excluded")
  ) %>%
  fmt_currency(columns = c(total_amount, median_amount, avg_amount), decimals = 0) %>%
  fmt_number(columns = n_payments, decimals = 0, sep_mark = ",") %>%
  fmt_percent(columns = pct_of_total, decimals = 1) %>%
  cols_label(
    payment_type_short = "Payment Type",
    n_payments = "Count",
    total_amount = "Total",
    median_amount = "Median",
    avg_amount = "Average",
    pct_of_total = "% of Total"
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(columns = total_amount, rows = pct_of_total > 0.15)
  ) %>%
  data_color(
    columns = pct_of_total,
    palette = c("#f8fafc", "#dbeafe", "#2563eb"),
    domain = c(0, max(company_summary$pct_of_total))
  ) %>%
  tab_source_note(source_note = "Source: CMS Open Payments · Auto-generated from pipeline") %>%
  tab_options(
    table.font.size = px(12),
    heading.title.font.size = px(15),
    column_labels.font.weight = "bold",
    table.border.top.style = "none",
    table_body.hlines.style = "none"
  )
```

## Where the money actually goes

The table is the deliverable, but the chart is what gets remembered. Same rules as the
other case studies: the title states the finding, labels sit on the data instead of a
legend, one idea per chart.

And the finding here is a good one. Count the payments and this dataset is a catering
ledger, nearly every record is a meal with an $18 median. Count the dollars and the
meals barely matter, the money is speaker fees and consulting.

```{r}
#| label: chart-style
#| code-summary: "Shared chart style"

library(ggplot2)

## same palette as the rest of the portfolio. navy carries the money story,
## slate fades the supporting series into the background
NAVY <- "#0f172a"; BLUE <- "#2563eb"; SLATE <- "#94a3b8"
GRAY <- "#64748b"; GREEN <- "#059669"; AMBER <- "#d97706"

theme_exec <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 14, color = NAVY),
      plot.subtitle = element_text(color = GRAY, size = 9.5),
      plot.caption = element_text(color = GRAY, size = 8.5),
      plot.title.position = "plot",
      panel.grid.minor = element_blank(),
      legend.position = "none"
    )
}
```

```{r}
#| label: count-vs-dollars
#| code-summary: "Share of payment count vs share of dollars by payment type"
#| fig-cap: "Share of payment records vs share of dollars by payment type, all manufacturers combined. The median meal is $18, the median speaker fee is $2,345."
#| fig-width: 8
#| fig-height: 4.6

## five buckets tell the story, everything else is rounding error
mix <- payments %>%
  mutate(bucket = case_when(
    payment_type_short %in% c("Speaker / services fees", "CME speaker fees") ~ "Speaker fees",
    payment_type_short %in% c("Meals", "Consulting", "Travel and lodging")   ~ payment_type_short,
    TRUE ~ "All other"
  )) %>%
  group_by(bucket) %>%
  summarise(n = n(), dollars = sum(amount_usd), .groups = "drop") %>%
  mutate(
    pct_n = n / sum(n) * 100,
    pct_d = dollars / sum(dollars) * 100
  )

## order by dollars so the money story reads top down
mix <- mix %>% arrange(pct_d)
mix$bucket <- factor(mix$bucket, levels = mix$bucket)

long <- mix %>%
  pivot_longer(c(pct_d, pct_n), names_to = "measure", values_to = "pct") %>%
  mutate(measure = factor(measure, levels = c("pct_n", "pct_d")))

### the top group's labels name the series, so no legend box needed
top_bucket <- levels(mix$bucket)[nrow(mix)]
long <- long %>%
  mutate(lab = case_when(
    bucket == top_bucket & measure == "pct_d" ~ sprintf("%.0f%% of the dollars", pct),
    bucket == top_bucket & measure == "pct_n" ~ sprintf("%.0f%% of the payments", pct),
    pct < 1   ~ sprintf("%.1f%%", pct),
    TRUE      ~ sprintf("%.0f%%", pct)
  ))

ggplot(long, aes(x = pct, y = bucket, fill = measure)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.62) +
  geom_text(aes(label = lab, color = measure),
            position = position_dodge(width = 0.72),
            hjust = -0.12, size = 3.4, fontface = "bold") +
  scale_fill_manual(values = c(pct_d = NAVY, pct_n = SLATE)) +
  scale_color_manual(values = c(pct_d = NAVY, pct_n = SLATE)) +
  scale_x_continuous(limits = c(0, 118), expand = c(0, 0)) +
  labs(
    title = "Almost every payment is a meal. Almost none of the money is",
    subtitle = sprintf("share of record count vs share of dollars by payment type, CMS program year %d, %d manufacturers", summary_year, n_parents),
    x = NULL, y = NULL
  ) +
  theme_exec() +
  theme(
    axis.text.x = element_blank(),       ## labels carry the values, axis can go
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10, color = NAVY)
  )
```

This is the chart a compliance officer or a journalist actually needs. Headlines about
pharma money usually count payments, which means they're mostly counting sandwiches.
The dollars sit somewhere else entirely.

## One company is half the money, every month

Roll the 26 filing names up to parents (Step 2) and the monthly cadence gets simple.
Without the rollup this chart is a lie, AbbVie's lead shrinks and Johnson & Johnson
splits into five entities that all look mid-pack.

```{r}
#| label: monthly-trend
#| code-summary: "Monthly payment volume by parent company"
#| fig-cap: "Monthly payments by parent company, top 4 of 7 shown, flagged outliers excluded. Names label the lines directly, no legend."
#| fig-width: 8
#| fig-height: 4.8

top4 <- payments %>%
  filter(!is_outlier) %>%
  count(parent, wt = amount_usd, name = "spend") %>%
  slice_max(spend, n = 4) %>%
  pull(parent)

monthly_trend <- payments %>%
  filter(!is_outlier, parent %in% top4) %>%
  group_by(year_month, parent) %>%
  summarise(total = sum(amount_usd) / 1e6, .groups = "drop")

trend_palette <- setNames(c(NAVY, BLUE, GREEN, AMBER), top4)

## abbvie's share of ALL dollars, the headline reads this
abbvie_share <- payments %>%
  summarise(s = sum(amount_usd[parent == top4[1]]) / sum(amount_usd) * 100) %>%
  pull(s)

## company names ride the end of their own lines. the bottom two finish the
## year close together so they each get a nudge, otherwise the labels collide
ends <- monthly_trend %>%
  group_by(parent) %>%
  slice_max(year_month, n = 1) %>%
  ungroup() %>%
  arrange(total) %>%
  mutate(label_y = total + c(-0.55, 0.55, 0, 0))

ggplot(monthly_trend, aes(year_month, total, color = parent)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_text(data = ends, aes(y = label_y, label = parent),
            hjust = 0, nudge_x = 9, size = 3.5, fontface = "bold") +
  scale_color_manual(values = trend_palette) +
  scale_y_continuous(labels = label_dollar(suffix = "M")) +
  ## explicit breaks, otherwise the room reserved for the line end labels
  ## grows phantom month ticks past december
  scale_x_date(breaks = seq(min(monthly_trend$year_month),
                            max(monthly_trend$year_month), by = "month"),
               date_labels = "%b",
               expand = expansion(mult = c(0.02, 0.24))) +
  labs(
    title = sprintf("%s alone is %.0f%% of all the %d money", top4[1], abbvie_share, summary_year),
    subtitle = "monthly payments by parent company, top 4 of 7, flagged outliers excluded",
    x = NULL, y = NULL
  ) +
  theme_exec() +
  theme(panel.grid.major.x = element_blank())
```

## Step 4: Render to Excel via `openxlsx2`

Some stakeholders need formatted Excel and pretending otherwise just costs you
credibility. Generate it from the same cleaned data, alongside the HTML, and everyone
gets the format they actually open.

```{r}
#| label: excel-export
#| code-summary: "Generate formatted Excel workbook"
#| eval: false

library(openxlsx2)

## three sheets: the summary, the line level detail, and the flagged
## outliers on their own sheet so nothing silently disappears
wb <- wb_workbook(creator = "Analytics Pipeline") %>%
  wb_add_worksheet("Summary", grid_lines = FALSE) %>%
  wb_add_data(x = company_summary, start_row = 3, start_col = 2) %>%
  wb_add_cell_style(dims = "B3:G3", horizontal = "center", text_bold = TRUE) %>%
  wb_freeze_pane(first_active_row = 4) %>%

  wb_add_worksheet("Detail", grid_lines = FALSE) %>%
  wb_add_data(
    x = payments %>% filter(parent == target_company, !is_outlier) %>%
      select(payment_date, payment_type_short, specialty, state, amount_usd),
    start_row = 1
  ) %>%
  wb_freeze_pane(first_active_row = 2) %>%

  wb_add_worksheet("Outliers Flagged", grid_lines = FALSE) %>%
  wb_add_data(
    x = payments %>% filter(parent == target_company, is_outlier),
    start_row = 1
  )

### in the real pipeline:
### wb_save(wb, paste0("output/", target_company, "_2024.xlsx"))
cat("in production: saved to output/", target_company, "_2024.xlsx\n", sep = "")
cat("workbook structure: 3 sheets (Summary, Detail, Outliers Flagged)\n")
```

## Step 5: The runner script

In production all of the above lives in functions called from a parameterized runner.
One command, full deliverable, any company:

```bash
# what the team actually runs each month
quarto render reports/monthly-summary.qmd \
  -P company:"AbbVie" \
  -P period:"2024" \
  --output-dir output/abbvie-2024/
```

Or loop the whole roster in one batch:

```r
## render reports for every parent company in one shot
companies_to_render <- unique(payments$parent)

purrr::walk(companies_to_render, function(co) {
  quarto::quarto_render(
    "reports/monthly-summary.qmd",
    execute_params = list(company = co, period = "2024"),
    output_file = paste0("output/", janitor::make_clean_names(co), "_2024.html")
  )
})
```

## Key Results

Every number below is computed when this page renders, nothing is hand typed.

::: {.metrics-row}

::: {.metric}
[`{r} fmt_n(nrow(payments))`]{.metric-value}
[Records processed]{.metric-label}
[Real CMS 2024, measured at render]{.metric-detail}
:::

::: {.metric}
[`{r} paste0(n_raw_names, " → ", n_parents)`]{.metric-value}
[Entity rollup]{.metric-label}
[Raw filing names to parent companies]{.metric-detail}
:::

::: {.metric}
[`{r} if (is.na(compression_ratio)) "n/a" else sprintf("%.0f×", compression_ratio)`]{.metric-value}
[Smaller on disk]{.metric-label}
[Parquet vs CSV, measured at render]{.metric-detail}
:::

::: {.metric}
[`{r} sprintf("%.1f%%", pct_outliers)`]{.metric-value}
[Outliers flagged]{.metric-label}
[3×IQR per payment type, flagged not removed]{.metric-detail}
:::

:::

## What this demonstrates

**R fluency where R is the right answer.** Quarto + gt + ggplot2 + openxlsx2 is a
genuinely first class reporting stack, and this shows the pieces working together the
way they would in production.

**Entity resolution is everywhere.** `{r} n_raw_names` raw payer names collapse to
`{r} n_parents` parent companies, and skipping that step changes the answer to every
company level question on this page. The rollup has rough edges worth admitting, the
regex map is maintained by hand, so the day CMS adds a new subsidiary it lands unmapped
until someone notices. The unmapped names keep their raw form on purpose so they show up
instead of vanishing into an Other bucket. Production wants a reviewed mapping table, not
a smarter regex.

**Flag, don't delete.** Outliers get flagged inside each payment type and carried
through to their own Excel sheet. A $22,000 royalty is normal and a $22,000 meal is not,
a global threshold can't tell those apart.

**The methodology paper, made concrete.** Everything here is the practical
implementation of [The Case for Code-Driven Reporting](../posts/code-driven-reporting.html),
the same discipline applied to a real recurring deliverable.

## What I'd extend in production

- **Orchestration** with [targets](https://docs.ropensci.org/targets/) for dependency
  tracking and incremental rebuilds
- **Tests** for the cleaning functions with `testthat`, which would catch upstream schema
  changes the day they happen, and a unit test on the company rollup map
- **renv lockfile** so the whole team renders against identical dependencies
- **CI/CD**, auto render on data refresh via GitHub Actions
- **Email and Slack delivery**, finished reports distributed without anyone asking

## Tools & Methods

::: {.skill-grid}
[R]{.skill .primary}
[Quarto]{.skill .primary}
[gt / gtExtras]{.skill .primary}
[ggplot2]{.skill .primary}
[openxlsx2]{.skill .primary}
[dplyr]{.skill}
[tidyr]{.skill}
[lubridate]{.skill}
[Entity Resolution]{.skill}
[Modular Function Design]{.skill}
[Parameterized Reports]{.skill}
:::
