---
title: "PATTERN TITLE — e.g., Reading a Geo Test Without Fooling Yourself"
description: "One-sentence hook shown on the listing card."
date: 2026-06-11
categories: [Experiment Readouts]
# To start an entry: copy this file into the matching category folder
# (trends/, comparison/, distributions/, experiments/, kpis/),
# rename it, set the category, and remove `draft: true`.
# image: optional — listing auto-uses the first rendered chart if omitted
draft: true
toc: true
---

::: {.section-label}
Analysis Playbook · CATEGORY
:::

## The business question

What does leadership actually ask? Frame the scenario in one or two sentences.

```{r}
#| label: synthetic-data
#| code-summary: "Show the data setup"
library(ggplot2)
library(dplyr)

set.seed(2026)
# Build the synthetic dataset that demonstrates this pattern.
# Engineer the exact effect the entry is about (trend break, mix shift, etc.).
```

## The wrong chart

Show the chart people usually reach for, and say what it hides or distorts.

```{r}
#| label: wrong-chart
#| fig-cap: "Why this misleads: ..."
# ggplot(...) ...
```

## The right chart

```{r}
#| label: right-chart
#| fig-cap: "What to look at: ..."
# ggplot(...) ...
```

## How to read it

The statistical read — what conclusion the chart supports, what it doesn't,
and the test or interval that backs it up.

## When to use it (and when not)

- **Use when:** ...
- **Avoid when:** ...
